Association between trauma severity models and opportunities for improvement: A retrospective cohort study

### 2022-09-15 studyprogressreport.Rmd
attempting to make intro longer

Abbreviations
=============
* ASA-PS - American Society of Anesthesiology Physical Status
* KUH - Karolinska University Hospital
* KUH TCQD - Karolinska University Hospital Trauma Care Quality Database
* KUH TR - Karolinska University Hospital Trauma Registry
* NORMIT - Norwegian Survival Prediction Model in Trauma
* OFI - Opportunities for improvement
* Ps - Probability of survival
* TRISS - Trauma and Injury Severity Score

@@

Management of severe trauma is highly time-sensitive and is also dependent on the actions of a multidisciplinary chain of healthcare providers. Errors in management are a common cause of preventable deaths [@prevdeathsghorbani; @prevdeathsmontmany; @prevdeathsteixeira]. Outcomes in severe trauma are dependent on the quality of care received, and it has been shown in the US that trauma quality improvement (QI) programs improve outcomes [@ustqi]. However, it has also been shown that the implementation of QI programs differs widely between clinics [@tqipanglo] and there is still progress to be made.

@@

The gold standard of trauma quality improvement programs worldwide are multidisciplinary morbidity and mortality (M&M) conferences.  M&M conferences review selected patient cases and attempt to establish whether there were opportunities for improvement (OFI) in each specific case, and, where necessary, implement corrective actions. The incidence of OFI function in this manner as a proxy for incidence of suboptimal care. The WHO guidelines for trauma quality improvement programmes recommend regularly scheduled multidisciplinary conferences and give specific recommendations as to their structure and format [@whoguidelines]. Despite these  recommendations, there is a still a great degree of variation amongst clinics in how M&M conferences are carried out, with a significant number of clinics still using an informal and unstructured format [@tqipanglo]. The WHO note themselves in their guidelines that M&M conferences are not used to their fullest potential globally [@whoguidelines].

The value of M&M conferences in long-term quality improvement is dependent on selection of appropriate cases for review. The process by which patient cases are selected for M&M review differs between clinics and case selection is often non-standardised [@mmreview; @tqipanglo]. This weakness of M&M review has been subject to considerable efforts to develop strategies to standardise case selection. The benefit of standardising case selection is twofold; it could increase identification of the cases that are most likely to have been mismanaged, and it could also save resources in the form of work hours. The latter benefit is not insignificant considering the large caseload found in most urban trauma centres. Some clinics have already made efforts to standardise case election; for example by using quality indicators, or audit filters, to flag cases for inclusion in M&M conferences [@tqipanglo]. Examples of quality indicators include time to first medical contact, adverse event rate, and mortality rate [@developqi]. The WHO recommends that all "deaths, complications, adverse events and errors" should be reviewed [@whoguidelines].

<!-- Explain the what and why of TSM -->

## Trauma severity scores
Trauma severity scores (TSS) are used in trauma quality improvement programs and research for their usefulness in benchmarking and in comparing trauma severity between patients [@whoguidelines]. The existence of these models is necessitated by the heterogeneous nature of trauma, due to the variety in its underlying causes, which affect both the type and severity of the injury. This heterogeneity poses challenges in comparing and evaluating trauma outcomes and has spurred the development of multiple TSS. TSS were initially intended for use in prehospital triage, and therefore were simplistic by design for ease of use [@traumareview], but over the years newer and more advanced models have been developed for quality assurance and research purposes. Today there are hundreds of TSS models. These models differ in their methodology but all have the aim of translating injury severity into quantitative values. However, an inherent weakness of any TSS model is that estimates may be unreliable when applied to a population different from the one the model was developed from. Trauma type and severity will differ between locales and trauma patient populations can be difficult to compare between disparate geographical regions. This weakness is one reason why new models continue to be developed and improved around the globe.

The Trauma and Injury Severity Score (TRISS), first developed in the US in 1987 [@triss] and most recently updated in 2010 [@triss2010], is the most commonly used model worldwide. TRISS is a combination index that incorporates patient age, the Injury Severity Score (ISS), the Revised Trauma Score (RTS), and whether the injury was blunt or penetrating [@triss]. The resulting value is a percentage that estimates the probability of survival (Ps). The ISS gives a score based on the three anatomical regions with the most severe injury. A limitation of the ISS is that it cannot take into account situations where one anatomical region has multiple injuries. The RTS is based on the physiological parameters GCS, systolic blood pressure, and respiratory rate. The RTS is difficult to apply in situations where the patient has already been intubated, which gives TRISS limited value in estimating Ps in certain populations.

In 2014 researchers in Norway released a new model, Norwegian Survival Prediction Model in Trauma (NORMIT), that attempts to address the weaknesses of TRISS and simultaneously provide a model based on Norwegian patient populations. NORMIT is based on the patient's age, New Injury Severity Score (NISS), comorbidities prior to traumatic injury based on the ASA physical status classification system (ASA-PS), and their RTS. NISS differs from ISS in that NISS allows for scoring of multiple injuries within the same anatomical region. NORMIT also provides a method to apply a GCS score for intubated patients when calculating RTS.

## Trauma severity scores as indicators for peer review

Some clinics already use TSS as audit filters when selecting cases for M&M review [@whoguidelines]. However, the extent to which TSS are associated with or can predict OFI as found in M&M conferences is not well understood. A study at this institution found that neither NORMIT nor TRISS were able to reliably discriminate between preventable and non-preventable deaths [@prevdeathsghorbani]. However, that study included only deaths and not other morbidities and thereby had a low sample size. It is not known whether predictive models have better performance when taking into account the entire population of trauma patients. If it could be found that an existing TSS model is associated with OFI, it could allow for another avenue of automatically identifying potentially mismanaged cases in a standardised fashion. The aim of this study is to evaluate how trauma severity models are associated with and to what extent they can predict opportunities for improvement.

@@

Participants include all patients registered in both the trauma registry and the TCQD at KUH. <!--När du kommit igång med analysen så ska faktiskt nummer in här (yes ska göra det) --> Furthermore, patients <15 years of age will be excluded as paediatric trauma is a seperate clinical entity.

@@

@@

The ISS is given by estimating the severity of physical injuries by anatomical region. Severity is assessed on a seven-point scale from "No injury" to "Unsurvivable injury". The value given is points on a scale from 0 to 75. The RTS is based on physiological parameters GCS, systolic blood pressure, and respiratory rate. Values are taken from arrival at hospital. These parameters are weighted and expressed in a score from 0 to 8, where a lower score indicates more severe injury.

@@

@@

Since all eligible participants must exist in both the trauma and the trauma care quality register, and the trauma care quality registry is a subset of the trauma registry, the number of participants is limited to the number of cases registered in the trauma care quality database. The trauma care quality database includes around 2200 patients who were selected for review between 2014 and 2021. The exact study size has not yet been established at the time of writing this report.

@@

The NORMIT and TRISS scores will be calculated using the following variables from the trauma registry database: ISS, NISS, RTS, GCS, systolic blood pressure, respiratory rate, ASA, and age.

@@

R, a programming language for statistical computing, will be used for compilation and analysis of data. Conversion and handling of variables will be carried out according to the SweTrau manual. This is a complete case study. In cases where data necessary for analysis is missing, i.e. the variables referred to above, the participant will be excluded from the study. Results will be presented with a confidence interval of 95%. A p-value of <0.05 will be considered significant.

@@

All patient information used in this study is anonymised and stored in a secure database. Patient medical records will not be accessed. Patients were not asked to give consent for inclusion in the KUH trauma registry. The ability and ethical approval to collect patient data for this registry without pre-emptively asking for patient consent is well-established, as collecting this data is deemed to be in the public interest according to article 6.1 e of GDPR regulations. Collecting data is in the public interest due to the fact that this patient register is used to improve management of trauma, which is a leading cause of death. Patients have the right to be removed from the register if they wish, and also have the right to a copy of data that is stored about them on request.

@@

>>>>

Abbreviations
=============
* ASA-PS - American Society of Anesthesiology Physical Status
* KUH - Karolinska University Hospital
* NORMIT - Norwegian Survival Prediction Model in Trauma
* OFI - Opportunities for improvement
* Ps - Probability of survival
* TCQD - Karolinska University Hospital Trauma Care Quality Database
* TRISS - Trauma and Injury Severity Score

@@

Management of severe trauma is highly time-sensitive and is also dependent on the actions of a multidisciplinary chain of healthcare providers. Errors in management are a common cause of preventable deaths [@prevdeathsghorbani; @prevdeathsmontmany; @prevdeathsteixeira]. Outcomes in severe trauma are dependent on the quality of care received, and it has been shown in the US that trauma quality improvement (QI) programs improve outcomes [@ustqi]. Multiple tools and interventions have been developed for use in trauma QI programs, including morbidity and mortality conferences, audit filters, and statistical analysis [@hakkenbrakprevdeath]. However, the implementation of QI programs differs widely between clinics [@tqipanglo] and there is still progress to be made.

@@

The gold standard of trauma QI programs worldwide are multidisciplinary morbidity and mortality (M&M) conferences. M&M conferences review selected patient cases, attempt to establish whether there were opportunities for improvement (OFI) in each specific case, and, where necessary, implement corrective actions. The occurrence of OFI function in this manner as a proxy for incidence of suboptimal care and preventable deaths. The WHO guidelines for trauma quality improvement programmes recommend regularly scheduled multidisciplinary conferences and give specific recommendations as to their structure and format [@whoguidelines]. Despite these  recommendations, there is a still a great degree of variation amongst clinics in how M&M conferences are carried out, with a significant number of clinics still using an informal and unstructured format [@tqipanglo]. The WHO note themselves in their guidelines that M&M conferences are not used to their fullest potential globally [@whoguidelines].

The value of M&M conferences in long-term quality improvement is largely dependent on selection of appropriate cases for review. The process by which patient cases are selected for M&M review differs between clinics and case selection is often non-standardised [@mmreview; @tqipanglo]. This weakness of M&M review has been subject to considerable efforts to develop strategies to standardise case selection. The benefit of standardising case selection is twofold; it could increase identification of the cases that are most likely to have been mismanaged, and it could also save resources in the form of work hours. The latter benefit is not insignificant considering the large caseload found in most urban trauma centres. The WHO recommends that all "deaths, complications, adverse events and errors" should be reviewed [@whoguidelines]. Some clinics have already made efforts to standardise case election; for example by using quality indicators, or audit filters, to flag cases for inclusion in M&M conferences [@tqipanglo]. Examples of quality indicators include time to first medical contact, adverse event rate, and mortality rate [@developqi]. Another potential quality indicator, that is already used at some clinics [@whoguidelines], are trauma severity scores.

## Trauma severity scores
Trauma as a clinical entity is highly heterogeneous due to the variety in its underlying causes, which affect both the type and severity of the injury. An important question for trauma QI programs to solve is how to evaluate and compare outcomes given the difficulties posed by the heterogeneity of trauma. This question has spurred the development of multiple trauma severity scores (TSS). TSS attempt to translate the severity of physical injury into a quantitative value by using physiological and anatomical parameters as predictors of mortality. TSS were initially intended for use in prehospital triage, and therefore were simplistic by design for ease of use [@traumareview], but over the years newer and more advanced algorithms have been developed to fulfil the needs of research, quality assurance, and clinical triage [@traumareview]. Today there are hundreds of TSS models with differing methodology and the common aim of expressing trauma severity as a numerical value that either directly or indirectly represents the probability of survival (Ps).

Trauma scoring has become a cornerstone of modern trauma QI programs [@historytss]. One example of their use is highlighting unexpectedly high mortality rates amongst patients with a high Ps [@whoguidelines]. TSS also allow clinics to compare their outcomes against an established benchmark [@whoguidelines]. However, an inherent weakness of any TSS model is that estimates may be unreliable when applied to a population different from the one the model was developed from. Trauma type and severity will differ between locales and trauma patient populations can be difficult to compare between disparate geographical regions [@assessingquality]. This weakness is one reason why models continue to be developed and improved on. Another reason is the efforts to establish which risk factors best predict mortality; different models use different physiological and anatomical parameters.

The Trauma and Injury Severity Score (TRISS), first developed in the US in 1987 [@triss] and most recently updated in 2010 [@triss2010], is the most commonly used model worldwide [@trissbetter]. TRISS is a combination index that incorporates patient age, the Injury Severity Score (ISS), the Revised Trauma Score (RTS), and whether the injury was blunt or penetrating [@triss]. The resulting value is a percentage that estimates the probability of survival (Ps). The ISS gives a score based on the three anatomical regions with the most severe injury. A limitation of the ISS is that it cannot take into account situations where one anatomical region has multiple injuries [@trissbetter]. The RTS is based on the physiological parameters GCS, systolic blood pressure, and respiratory rate. The GCS cannot be estimated in situations where the patient has already been intubated, since a valid verbal score can be ascertained, which gives TRISS limited value in estimating Ps in certain populations [@trissbetter].

In 2014 researchers in Norway released a new model, the Norwegian Survival Prediction Model in Trauma (NORMIT), that attempts to address the weaknesses of TRISS and simultaneously provide a model based on Norwegian patient populations [@normit]. NORMIT is based on the patient's age, New Injury Severity Score (NISS), comorbidities prior to traumatic injury based on the ASA physical status classification system (ASA-PS), and their RTS. NISS differs from ISS in that NISS allows for scoring of multiple injuries within the same anatomical region. Validation of NISS has shown that NISS is better at predicting mortality in trauma patients compared to the ISS, especially in patients with multiple severe injuries within one anatomical region [@nissbetter]. NORMIT also provides a method to incorporate the GCS for intubated patients when calculating RTS. For patients who arrive at hospital and under general anaesthesia, the RTS is calculated based on the prehospital value prior to intubation [@normit]. NORMIT was updated in 2018 [@skagaValidatingPerformanceTRISS2018] and a validation study at this institution found that the updated NORMIT reliably predicted survival at a Swedish trauma centre but both the old and new versions of NORMIT performed worse when applied to the entire national population [@validationnormit]

## Trauma severity scores as quality indicators

As already mentioned, some clinics use TSS as audit filters to aid in identifying cases for M&M review. However, the extent to which TSS are associated with or can predict OFI as found in M&M conferences is not well understood. A study at this institution found that neither NORMIT nor TRISS were able to reliably discriminate between preventable and non-preventable deaths [@prevdeathsghorbani]. However, that study included only deaths and not other morbidities and thereby had a low sample size. It is not known whether predictive models have better performance when taking into account the entire population of trauma patients.

If it could be found that an existing TSS model is associated with OFI, it could allow for another avenue of automatically identifying potentially mismanaged cases in a standardised fashion. Conversely, if TSS models are not associated with OFI, using TSS as audit filters may result in cases being unnecessarily flagged for review and in doing so waste conference resources.

There are far more TSS models than the two described above. These two models are of interest for specific reasons: TRISS is the most widely used model and NORMIT is based on a Scandinavian population, which is the location of this institution. It seems reasonable to begin by investigating these two models. The aim of this study is to evaluate how the trauma severity models TRISS and NORMIT are associated with and to what extent they can predict opportunities for improvement.

<!-- It is possible that I will add other scores later, but in the interest of having this document done in time to submit to school, I'm going with TRISS & NORMIT for now. I mention under the "development and time plan" heading that I may add scores later -->

@@

Participants include all patients registered in both the trauma registry and the TCQD at KUH. <!--När du kommit igång med analysen så ska faktiskt nummer in här (yes ska göra det) -->

@@

### Study-specific exclusion criteria
Patients <15 years of age will be excluded as paediatric trauma is a separate clinical entity. Furthermore, this is a complete case study and all patients will be excluded where there is not enough data recorded in order to carry out the analysis.

@@

The ISS is given by estimating the severity of physical injuries by anatomical region. Severity is assessed on a seven-point scale from "No injury" to "Unsurvivable injury". The values from the three anatomical regions with the most severe injury are combined into a score that is given is points on a scale from 0 to 75. The RTS is based on physiological parameters GCS, systolic blood pressure, and respiratory rate. Values are taken from arrival at hospital. These parameters are weighted and expressed in a score from 0 to 8, where a lower score indicates more severe injury.

@@

NISS is based on the same variables as ISS. It differs from ISS in that it takes into account multiple injuries within one anatomical region and calculates a score based on the three most severe injuries regardless of anatomical region.

@@

Since all eligible participants must exist in both the trauma registry and the TCQD, and the TCQD is a subset of the trauma registry, the maximum number of participants is limited to the number of cases registered in the TCQD. The TCQD includes around 2200 patients who were selected for review between 2014 and 2021. In order to establish the exact study size, the eligibility criteria need to first be applied. The exact study size has not yet been established at the time of writing this report.

@@

The NORMIT and TRISS scores will be calculated using the following variables from the trauma registry database: ISS, NISS, RTS, GCS, systolic blood pressure, respiratory rate, ASA, and age. The scores will be calculated according to the formulas given by the respective developers of each score.

@@

R, a programming language for statistical computing, will be used for compilation and analysis of data. Conversion and handling of variables will be carried out according to the SweTrau manual. In cases where data necessary for analysis is missing, i.e. the variables referred to above, the participant will be excluded from the study. Logistic regression will be used to estimate the predictive value of each TSS in regards to OFI. Results will be presented with a confidence interval of 95%. A p-value of <0.05 will be considered significant.

@@

All patient information used in this study is anonymised and stored in a secure database. Patient medical records will not be accessed. Patients were not asked to give consent for inclusion in the KUH trauma registry. The ability and ethical approval to collect patient data for this registry without pre-emptively asking for patient consent is well-established, as collecting this data is deemed to be in the public interest according to article 6.1 e of GDPR regulations. Collecting data is in the public interest due to the fact that this patient register is used to improve management of trauma, which is a leading cause of death. Patients have the right to be removed from the register on request. They also have the right to a copy of data that is stored about them on request.

@@

It is possible that scores other than NORMIT and TRISS will be analysed if there is time and it is feasible based on the data I have access to. I will decide this within the first week or two of starting my data analysis code.
